---
title: "RMS Titanic"
author: "Isabella Abilheira"
date: Feb 1 2023
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|-----------------|--------------------------|-----------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look

<!-- -------------------------------------------------- -->

### **q1** Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
df_titanic %>% glimpse()
df_titanic
```

**Observations**:

-   Class: 1st, 2nd, 3rd, Crew
-   Sex: Male, Female
-   Age: Child, Adult
-   Survived: Yes, No
-   n: count for each unique combination of the other 4 variables

### **q2** Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

-   Are there any differences?
    -   Yes there is a small difference, Wikipedia states the total number of passengers is 2224 and our data set states the total is 2201
-   If yes, what might account for those differences?
    -   (In)accuracy of ticketing documentation
    -   Differences between ticketing and boarding documentation
        -   People who cancelled their trip last minute
        -   Passengers who traveled under aliases

### **q3** Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
survived <- df_titanic %>% filter(Survived == "Yes")

survived %>%
  ggplot(aes(x = Class, y = n, fill = Sex)) +
  geom_col()
```

```{r q3-numeric-validation}
crew <- df_titanic %>% filter(Class == "Crew")
crew

women3 <- df_titanic %>% filter(Class == "3rd", Sex == "Female", Survived== "Yes")
women3

men3 <- df_titanic %>% filter(Class == "3rd", Sex == "Male", Survived== "Yes")
men3
#crew %>% summarize(total = sum(n))
```

**Observations**:

-   More women in first class survived compared to men in first class
    -   More women survived in first class than any other class
    -   In 1st, 2nd, and 3rd class more women survived than men
        -   This validates the women part of the "women and children first" protocol that was followed when loading the life boats
-   The most men who survived overall were from the Crew
    -   Important to note there were more males in the crew than females

# Deeper Look

<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### **q4** Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
survived_prop <- df_prop %>% filter(Survived == "Yes")
survived_prop

survived_prop %>%
  ggplot(aes(x = Class, y = Prop, fill = Sex)) +
  geom_col()
```

**Observations**:

-   According to this graph, the proportion of women surviving compared to men is higher for every class

**Is there anything *fishy* going on in your plot?**

-   Yes, the graph is inaccurately stating that over 100% of the passengers in each class survived

### **q5** Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}

survived_prop %>%
  ggplot(aes(x = Class, y = Prop)) +
  geom_col() +
  facet_grid(vars(Sex), vars(Age))
  
```

**Observations**:

-   All children in 1st and 2nd class survived
-   No children in Crew survived (since there were no children in the Crew)
-   A higher proportion of female crew survived compared to male
    -   Important to note there were more males in the crew than females

**If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.**

-   By analyzing each variable individually and then comparing them against each other allows us to have a more accurate and detailed representation of the data

    -   For example, with this new plot, we are able to see that over 90% of female Adults, 100% of both female and male children and over 25% of male adults in 1st class survived. The other plot only inaccurately portrayed that over 100% of all males in 1st class survived and over 150% of all females in 1st class survived.

# Notes

<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
