---
title: "Gapminder"
author: "Isabella Abilheira"
date: 2/22/23
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|-------------------|----------------------------|-------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an independent organization that seeks to educate people about the state of the world. They seek to counteract the worldview constructed by a hype-driven media cycle, and promote a "fact-based worldview" by focusing on data. The dataset we'll study in this challenge is from Gapminder.

# Guided EDA

<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### **q0** Perform your "first checks" on the dataset. What variables are in this

dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
gapminder %>% glimpse
```

**Observations**:

-   6 variables: country, continent, year, lifeExp, pop, gdpPercap

### **q1** Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- max(gapminder %>% pull(year))

year_min <- min(gapminder %>% pull(year))
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### **q2** Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent

gapminder %>%
  filter(year == year_min) %>% 
    ggplot() +
    stat_boxplot(aes(continent, gdpPercap)) +
    scale_y_log10() +
    coord_flip()

```

**Observations**:

-   Each of Africa's gdpPercap were less than 5000
-   Oceania's gdpPercap were both around 10000
-   Asia's gdpPercap were all less than 10000
-   The Americas' gdpPercap has a cluster around 2500
-   Europe has the second highest gdpPercap at almost 150000
-   Asia had one gdpPercap that was over 90000 which is not included in this visual

**Difficulties & Approaches**:

-   Asia had one gdpPercap that was greater than 90000 which was an outlier and squished the rest of the data so I removed it for clarity
-   Oceania's box was unreadable vertically so i flipped the graph

### **q3** You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2
gapminder %>%
  filter(year == year_min) %>% 
  group_by(continent) %>% 
  filter(!continent == "Oceania") %>% 
  filter(gdpPercap < quantile(gdpPercap, 0.01) | gdpPercap > quantile(gdpPercap, 0.99))
```

**Observations**:

-   Identify the outlier countries from q2
    -   Asia: gdpPercap 331.0000 and 108382.3529
    -   Europe: gdpPercap 973.5332 and 14734.2327
    -   Africa: gdpPercap 298.8462 and 4725.2955
    -   Americas: gdpPercap 1397.7171 and 13990.4821

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )

```

### **q4** Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder %>%
  filter(year == max(year) | year == min(year)) %>%
  
  ggplot(aes(continent, gdpPercap)) +
  geom_boxplot() +
  scale_y_log10() +
  facet_wrap(~year) +
  
  geom_point(
    data = . %>% 
      group_by(year) %>%
      filter(
          gdpPercap < quantile(gdpPercap, 0.01) | 
          gdpPercap > quantile(gdpPercap, 0.99)
      ),
    mapping = aes(color = country),
    size = 2
  ) +
  geom_label_repel(
    data = . %>%
      filter(
        country %in% c("Guinea-Bissau", "Liberia"),
        year == 1952
      ),
    mapping = aes(label = country)
  )
```

**Observations**:

-   Kuwait is an outlier in both 1952 and 2007
-   All of the countries gdpPercap median increased from 1952 to 2007
-   Asia's, Africa's, America's and Europe's upper quartile increased significantly from 1952 to 2007
    -   Europe's upper quartile had the largest increased
    -   Additionally, Europe's lower quartile increased the most from 1952 to 2007
-   Oceania's data is still hard to read likely because there are fewer data points

# Your Own EDA

<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### **q5** Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
## TASK: Your first graph
## population and lifeExp
gapminder %>%
  filter(year == max(year) | year == min(year)) %>%
           
  ggplot(aes(lifeExp, pop)) +
  geom_point(
    mapping = aes(color = continent),
    size = 2
  ) +
  scale_y_log10() +
  facet_grid(vars(continent), vars(year))

```

-   lifeExp of all continents seem to have generally gone up from 1952 to 2007
-   pop of all continents seem to have generally gone up from 1952 to 2007
-   All countries in Europe's lifeExp in 2007 is above 70 years
-   Asia in 2007 has one major outlier where the lifeExp is around 45 years
    -   the rest of Asia's life expectancies span from approx. 60-85 years
-   The America's lifeExp had quite a wide range in 1952, which decreased by 2007
    -   Africa had the opposite trend where there was a small span in 1952 and the span increased by 2007
-   Asia was the continent with the lowest lifeExp in 1952
    -   Africa was the continent with the lowest lifeExp in 2007

```{r q5-task2}
## TASK: Your second graph
## population and gdp

gapminder %>%
  filter(year == max(year)) %>%
           
  ggplot(aes(pop, gdpPercap)) +
  geom_point(
    mapping = aes(color = continent),
    size = 2
  ) +
  scale_x_log10() +
  facet_grid(vars(year), vars(continent))
```

-   (Your notes and observations here)

```{r q5-task3}
## TASK: Your third graph
```

-   (Your notes and observations here)
